---
title: "Ongava Climate Analysis"
author: "Heather Throop"
format: html
editor: visual
editor_options: 
  chunk_output_type: console
---

This code explores historical precipitation data for Ongava Research Center and uses this to come up with a watering schedule for the decomposition experiment establish by Heather Throop and Rolf Becker in June 2024.

Rainfall data from the experiment are compared with historical data.

Data are read in from Dropbox, with the exception of WorldClim geotiffs that are pulled from a folder on Heather's local drive (due to issues with getting an API pull to work with Dropbox - see notes below).\
Output csv data are saved in Dropbox.\
Output figures are saved in the Github repo (along with this script).

# Set Up

```{r}
#| label: setup

# Load necessary packages
library(terra)
library(here)
library(tidyverse)
library(curl)
library(lubridate)

# Define the coordinates for Ongava Research Centre
lat <- -19.35472
lon <- 15.90289

# define the dropbox folder path where output will be saved
dropbox_path <- "/Users/hthroop/ASU Dropbox/Heather Throop/DERT (ThroopLab) ASU/Lab Data & Metadata Archives/OngavaDungDecay/RainfallData/R_output"
```

# WorldClim Data

## 2010-2019 WorldClim Monthly PPT Data

It should be possible to pull data from UC Davis site but I have not been successful with this (looks like a web issue rather than coding issue). Here instead I download the geotiffs for 2010-2019 only.

```{r}

# Define the time range (this could be for a longer time, but only downloaded the 2010-2019 files while at Ongava due to slow connectivtity - had to download for whole world dut to server access issues)
start_year <- 2010
end_year <- 2019

# Path to the directory where the downloaded .tif files are stored
# data_dir <- here("wc2.1_cruts4.06_2.5m_prec_2010-2019")
# Currently this maps to Heather's computer, local version of the Dropbox data file (I can't get an API working to Dropbox to do a data pull)
data_dir <- file.path("/Users/hthroop/ASU Dropbox/Heather Throop/DERT (ThroopLab) ASU/Lab Data & Metadata Archives/OngavaDungDecay/wc2.1_cruts4.06_2.5m_prec_2010-2019")

# Function to extract WorldClim2.1 data from local files
get_WorldClim2_data <- function(lat, lon, start_year, end_year, data_dir) {
  # Create an empty data frame to store the results
  precip_data <- data.frame(Year = integer(), 
                            Month = integer(), 
                            Precipitation = numeric())
  
  # Convert coordinates to matrix
  coordinates <- matrix(c(lon, lat), ncol = 2)
  
  # Loop through each year and month and process the corresponding data
  for (year in start_year:end_year) {
    for (month in 1:12) {
      # Construct the file path for the specific year and month
      file_path <- file.path(data_dir, sprintf("wc2.1_2.5m_prec_%d-%02d.tif", year, month))
      
      # Debugging step: print the file path being checked
      cat("Checking file:", file_path, "\n")
      
      # Check if the file exists before attempting to load it
      if (file.exists(file_path)) {
        cat("File exists: ", file_path, "\n")
        
        # Load the raster file using terra
        raster_data <- rast(file_path)
        
        # Extract the precipitation data for the location
        precip_value <- terra::extract(raster_data, coordinates)
        
        # Check if precip_value has the expected structure and contents
        if (!is.null(precip_value) && nrow(precip_value) > 0) {
          precip_value <- precip_value[1, 1]  # Extract the value from the first row and first column
          precip_data <- rbind(precip_data, data.frame(Year = year, Month = month, Precipitation = precip_value))
        } else {
          precip_data <- rbind(precip_data, data.frame(Year = year, Month = month, Precipitation = NA))
        }
        
      } else {
        warning(paste("File not found:", file_path))
        # Append NA for missing file
        precip_data <- rbind(precip_data, data.frame(Year = year, Month = month, Precipitation = NA))
      }
    }
  }
  
  return(precip_data)
}

# Get WorldClim2.1 data for the specified location and date range
precip_data_2010.2019 <- get_WorldClim2_data(lat, lon, start_year, end_year, data_dir)

# Display the summary of the data
print(summary(precip_data_2010.2019))

# Save the precip_data dataframe as a CSV file to Dropbox
write.csv(precip_data_2010.2019, file = file.path(dropbox_path, "precip_data_2010.2019.csv"), 
          row.names = FALSE)

```

### PPT Summaries

#### Monthly PPT (as percentage of annual total)

```{r}
# Calculate for each month the percentage of the total PPT for that year that fell in that month

# Function to calculate the percentage of monthly precipitation out of the annual total
calculate_pct_annual_PPT <- function(precip_data) {
  # Group data by year and calculate the annual total precipitation
  annual_totals <- precip_data |>
    group_by(Year) |>
    summarise(Annual_PPT = sum(Precipitation, na.rm = TRUE))
  
  # Join the annual totals back to the original data
  precip_data <- precip_data |>
    left_join(annual_totals, by = "Year")
  
  # Calculate the percentage of each month's precipitation out of the annual total
  precip_data <- precip_data |>
    mutate(pct_annual_PPT = (Precipitation / Annual_PPT) * 100)
  
  return(precip_data)
}

# For the WorldClim data at Ongava, calculate for each month the percentage of the total PPT for that year that fell in that month

# Apply the function created above to do this, creating two new variables:
# Annual_PPT 
# pct_annual_PPT
precip_data_2010.2019 <- calculate_pct_annual_PPT(precip_data_2010.2019)
```

#### Summarize monthly PPT (2010-2019)

```{r}
# Summarize mean monthly PPT for WorldClim data   
monthly_precip_2010.2019 <- precip_data_2010.2019 |>
    group_by(Month) |>
    summarise(mean_monthly_PPT = mean(Precipitation, na.rm = TRUE),
              N = n(),
              sd_monthly_PPT = sd(Precipitation),
              se_monthly_PPT = sd_monthly_PPT / sqrt(N))
 
```

##### Plot

```{r}
# plot monthly PPT patterns from WorldClim2.1 data
monthly_rainfall_plot <- ggplot(monthly_precip_2010.2019, aes(x = Month)) +
  geom_line(aes(y = mean_monthly_PPT), linewidth = 1) +
  geom_errorbar(aes(ymin = mean_monthly_PPT - se_monthly_PPT, 
                    ymax = mean_monthly_PPT + se_monthly_PPT), width = 0.2) +
  scale_x_continuous(breaks = 1:12) +
  scale_y_continuous(expand = c(0, 0)) +
  labs(x = "Month", y = "Precipitation (mm)", 
       title = "Ongava WorldClim2.1, 2009-2019") +
  theme_classic() 
monthly_rainfall_plot

ggsave(filename = here("figs", "2009-2019monthly_rainfall_plot.jpg"),
       plot = monthly_rainfall_plot,
       device = "jpeg",
       width = 8, height = 6, dpi = 300)
```

#### Rain year summary

```{r}
# Add the rain_year column to precip_data
precip_data_2010.2019 <- precip_data_2010.2019 |>
  mutate(rain_year = ifelse(Month >= 9 & Month <= 12, 
                            paste0(Year, "-", Year + 1),
                            paste0(Year - 1, "-", Year)))

```

##### NOT WORKING: monthly PPT (as percentage of rainyear total)

```{r}

# Function to calculate the percentage of monthly precipitation out of the annual rainyear total
calculate_pct_annual_rainyr_PPT <- function(precip_data) {
  # Group data by year and calculate the annual total precipitation
  annual_totals <- precip_data_2010.2019 |>
    group_by(rain_year) |>
    summarise(Annual_PPT = sum(Precipitation, na.rm = TRUE))
  
  # Join the annual totals back to the original data
  precip_data <- precip_data |>
    left_join(annual_totals, by = "rain_year")
  
  # Calculate the percentage of each month's precipitation out of the annual total
  precip_data <- precip_data |>
    mutate(pct_annual_PPT = (Precipitation / Annual_PPT) * 100)
  
  return(precip_data)
}

# Apply the function to calculate monthly percent of total yearly rainfall the precip_data dataframe
precip_data_2010.2019 <- calculate_pct_annual_rainyr_PPT(precip_data_2010.2019)

```

## 1960-2021 WorldClim Data

It should be possible to pull data from UC Davis site but I have not been successful with this (looks like a web issue rather than coding issue). The alternative is to download the ENTIRE dataset (16 GB). I did this but extracted the monthly precip data from the geotiffs and deleted the geotiffs.

### Download WorldClim Data - DO NOT RUN (USE DOWNLOADED DATA)

DO NOT RUN THIS CHUNK – PULLS DATA FROM GEOTIFFS (NOW DELETED)

If needed, these can be redownloaded from <https://worldclim.org/data/monthlywth.html> and placed in the (empty) folder referenced in this chunk.

```{r}
#### DO NOT RUN THIS CHUNK - USES DELETED WORLDCLIM DATA
# Define the time range
#start_year <- 1960
#end_year <- 2021

# Path to the directory where the downloaded .tif files are stored
#data_dir <- here("wc2.1_cruts4.06_2.5m_prec_1960-2021")

# Function to extract WorldClim2.1 data from local files
#get_WorldClim2_data <- function(lat, lon, start_year, end_year, data_dir) {
  # Create an empty data frame to store the results
 # precip_data <- data.frame(Year = integer(), 
 #                           Month = integer(), 
 #                           Precipitation = numeric())
  
  # Convert coordinates to a matrix
 # coordinates <- matrix(c(lon, lat), ncol = 2)
  
  # Loop through each year and month and process the corresponding data
#  for (year in start_year:end_year) {
#    for (month in 1:12) {
      # Construct the file path for the specific year and month
    #  file_path <- file.path(data_dir, sprintf("wc2.1_2.5m_prec_%d-%02d.tif", year, month))
      
      # Check if the file exists before attempting to load it
#      if (file.exists(file_path)) {
        # Load the raster file using terra
 #       raster_data <- rast(file_path)
#        
        # Ensure the raster has the correct CRS (WGS84)
 #       crs(raster_data) <- "EPSG:4326"  # WGS84
        
        # Extract the precipitation data for the location
#        precip_value <- terra::extract(raster_data, coordinates)
        
        # Check if precip_value has the expected structure and contents
 #       if (!is.null(precip_value) && nrow(precip_value) > 0) {
   #       precip_value <- precip_value[1, 1]  # Extract the value from the first row and first column
 #         precip_data <- rbind(precip_data, data.frame(Year = year, Month = month, Precipitation = precip_value))
#        } else {
#          precip_data <- rbind(precip_data, data.frame(Year = year, Month = month, Precipitation = NA))
 #       }
  #    } else {
 #       warning(paste("File not found:", file_path))
        # Append NA for missing file
#        precip_data <- rbind(precip_data, data.frame(Year = year, Month = month, Precipitation = NA))
      }
    }
  }
  
#  return(precip_data)
}

# Get WorldClim2.1 data for the specified location and date range
#precip_data_1960.2021 <- get_WorldClim2_data(latitude, longitude, start_year, end_year, data_dir)

# Display the summary of the data
#print(summary(precip_data_1960.2021))

# Save the precip_data dataframe as a CSV file
# Commented out just in case this chunk run accidentally
# write.csv(precip_data_1960.2021, file = here(
#  "output", "precip_data_1960.2021.csv"), row.names = FALSE))
 
```

### PPT Summaries (using saved data)

The following chunks use the saved data - ok to run!

```{r}

# Pull in the saved data extracted from WorldClim (in last chunk)
precip_data_1960.2021 <- read_csv(file = here(
  "output", "precip_data_1960.2021.csv"))

# Add the rain_year column to precip_data
precip_data_1960.2021 <- precip_data_1960.2021 |>
  mutate(rain_year = ifelse(Month >= 9 & Month <= 12, 
                            paste0(Year, "-", Year + 1),
                            paste0(Year - 1, "-", Year)))

# For the WorldClim data at Ongava, calculate for each month the percentage of the total PPT for that year that fell in that month. Use the function from above.
# Apply the function to the precip_data dataframe
precip_data_1960.2021 <- calculate_pct_annual_PPT(precip_data_1960.2021)
```

#### Summarize monthly PPT (1960-2021)

```{r}
# Summarize mean monthly PPT for WorldClim data   
monthly_precip_1960.2021 <- precip_data_1960.2021 |>
    group_by(Month) |>
    summarise(mean_monthly_PPT = mean(Precipitation, na.rm = TRUE),
              N = n(),
              sd_monthly_PPT = sd(Precipitation),
              se_monthly_PPT = sd_monthly_PPT / sqrt(N))
 
```

```{r}
# plot monthly PPT patterns from WorldClim2.1 data
monthly_rainfall_plot_1960.2021 <- ggplot(
    monthly_precip_1960.2021, aes(x = Month)) +
  geom_line(aes(y = mean_monthly_PPT), linewidth = 1) +
  geom_errorbar(aes(ymin = mean_monthly_PPT - se_monthly_PPT, 
                    ymax = mean_monthly_PPT + se_monthly_PPT), width = 0.2) +
  scale_x_continuous(breaks = 1:12) +
  scale_y_continuous(expand = c(0, 0)) +
  labs(x = "Month", y = "Precipitation (mm)", 
       title = "Ongava WorldClim2.1, 1960-2021") +
  theme_classic() 
monthly_rainfall_plot_1960.2021

ggsave(filename = here("figs", "1960-2021monthly_rainfall_plot.jpg"),
       plot = monthly_rainfall_plot_1960.2021,
       device = "jpeg",
       width = 8, height = 6, dpi = 300)
```

#### NOT WORKING - Rain year summary

```{r}
# Add the rain_year column to precip_data
precip_data_1960.2021 <- precip_data_1960.2021 |>
  mutate(rain_year = ifelse(Month >= 9 & Month <= 12, 
                            paste0(Year, "-", Year + 1),
                            paste0(Year - 1, "-", Year)))


### NOT WORKING AFTER HERE
# Apply the function to the precip_data dataframe
precip_data_1960.2021 <- calculate_pct_annual_rainyr_PPT_1960.2021(precip_data_1960.2021)

# Apply the function to calculate monthly percent of total yearly rainfall the precip_data dataframe
precip_data_1960.2021 <- calculate_pct_annual_rainyr_PPT(precip_data_1960.2021)

```

##### Plot

```{r}
# Plot WorldClim rain_year data
precip_data_1960.2021$rain_year <- 
  factor(precip_data_1960.2021$rain_year, 
         levels = unique(precip_data_1960.2021$rain_year))

# Remove monthly data
annual_rainfall_1960.2021 <- precip_data_1960.2021 |>
    filter(Month == 1) |>
    select(rain_year, Annual_rainyr_PPT)

# calculate mean rain_year PPT for the duration of the dataset
mean_rainyr_ppt <- mean(annual_rainfall_1960.2021$Annual_rainyr_PPT, 
                        na.rm = TRUE)

# make the plot
rainfall_1960.2021 <- ggplot(annual_rainfall_1960.2021, 
        aes(x = rain_year, y = Annual_rainyr_PPT, group = 1)) +
  geom_line(linewidth = 1) +
  labs(x = "Rain Year", y = "Precipitation (mm)", 
       title = "WorldClim2.1: Ongava Research Centre") +
  geom_line(linewidth = 1) +
  geom_hline(yintercept = mean_rainyr_ppt, color = "red", 
             linetype = "dashed", size = 1) + # line for mean rainfall PPT 
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "top") +
   scale_x_discrete(breaks = levels
                    (annual_rainfall_1960.2021$rain_year)[seq(1, length(levels(annual_rainfall_1960.2021$rain_year)), by = 5)]) 
rainfall_1960.2021

ggsave(filename = here("figs", "1960.2021rainyear_PPT_plot.jpg"),
       plot = rainfall_1960.2021,
       device = "jpeg",
       width = 8, height = 6, dpi = 300)
```

##### Rainyear histogram

```{r}
rainfall_histogram <- ggplot(annual_rainfall_1960.2021, aes(x = Annual_rainyr_PPT)) +
  geom_histogram(binwidth = 10, fill = "steelblue", color = "black") +
  labs(x = "Annual Rain Year Precipitation (mm)", y = "Count of Rain Years", 
       title = "Distribution of Rain Years by Annual Precipitation") +
  theme_classic()

rainfall_histogram
```

### Extreme Rainfall Years - NOT WORKING

```{r}
# Explore patterns of extreme rainfall years through time

# Calculate percentiles for: 
# "normal rainfall" (0.45 and 0.55 percentiles)
# "extreme low rainfall (0.10 percentile)
# "extremely high rainfall" (0.90) percentile. 
# These definitons are based on Knapp et al. 2015

percentiles <- quantile(annual_rainfall_1960.2021$Annual_rainyr_PPT, probs = c(0.10, 0.45, 0.55, 0.90), na.rm = TRUE)
```

#### Plot

```{r}
# Create a plot with horizontal lines for the mean and percentiles
rainfall_extremes_1960.2021 <- ggplot(annual_rainfall_1960.2021, 
      aes(x = rain_year, y = Annual_rainyr_PPT, group = 1)) +
  geom_line(linewidth = 1) +
  # Add the mean line
  geom_hline(yintercept = mean(
    annual_rainfall_1960.2021$Annual_rainyr_PPT, na.rm = TRUE), 
             color = "red", linetype = "dashed", size = 1) +
  # Add the percentile lines
  geom_hline(yintercept = percentiles[1], color = "blue", 
             linetype = "dotted", size = 1) + # 10th percentile
  geom_hline(yintercept = percentiles[2], color = "darkgreen", 
             linetype = "dotted", size = 1) + # 45th percentile
  geom_hline(yintercept = percentiles[3], color = "darkgreen", 
             linetype = "dotted", size = 1) + # 55th percentile
  geom_hline(yintercept = percentiles[4], color = "blue", 
             linetype = "dotted", size = 1) + # 90th percentile
  labs(x = "Rain Year", y = "Precipitation (mm)", 
       title = "1960-2021 WorldClim2.1 w average and extreme rainfall") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "top") +
  scale_x_discrete(breaks = levels(
    annual_rainfall_1960.2021$rain_year)[seq(1, length(
      levels(annual_rainfall_1960.2021$rain_year)), by = 5)]) 
rainfall_extremes_1960.2021

ggsave(filename = here("figs", "1960.2021rainfall_extremes.jpg"),
       plot = rainfall_1960.2021,
       device = "jpeg",
       width = 8, height = 6, dpi = 300)

```

## WorldClim Comparison (2010-2019 vs 1960-2021)

```{r}
# Comparison of recent (2010-2019) and longer-term (1960-2021) WorldClim data on monthly rainfall. This is to assess if there is an apparent change in distribution.

# summarize monthly PPT means for 2010-2019 WorldClim
monthly_precip_2010.2019 <- precip_data_2010.2019 |>
    group_by(Month) |>
    summarise(n = n(),
              mean_monthly_PPT = mean(Precipitation),
              sd_monthly_PPT = sd(Precipitation),
              se_monthly_PPT = sd_monthly_PPT/sqrt(n),
              mean_pct_annual_PPT = mean(pct_annual_PPT))

# summarize monthly PPT means for 1960-2021 WorldClim data 
monthly_precip_1960.2021 <- precip_data_1960.2021 |>
    group_by(Month) |>
    summarise(n = n(),
              mean_monthly_PPT = mean(Precipitation),
              sd_monthly_PPT = sd(Precipitation),
              se_monthly_PPT = sd_monthly_PPT/sqrt(n),
              mean_pct_annual_PPT = mean(pct_annual_PPT))

# Now compare these two time periods visually
# Add a column to each dataset indicating the time period
monthly_precip_2010.2019 <- monthly_precip_2010.2019 |>
  mutate(Period = "2010-2019")

monthly_precip_1960.2021 <- monthly_precip_1960.2021 |>
  mutate(Period = "1960-2021")

# Combine the two datasets into one
combined_WC_precip <- bind_rows(monthly_precip_2010.2019, monthly_precip_1960.2021)

# Plot the comparison
comparison_plot <- ggplot(combined_WC_precip, aes(x = Month, y = mean_monthly_PPT, color = Period, group = Period)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  labs(x = "Month", y = "Mean Monthly Precipitation (mm)", 
       title = "Comparison of Mean Monthly Precipitation (2010-2019 vs 1960-2021)",
       color = "Period") +
  theme_minimal() +
  scale_x_continuous(breaks = 1:12, labels = month.abb) +  # Convert month numbers to abbreviated month names
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
comparison_plot
# There is not much difference in the distribution for the two time periods. I will use the 1960-2021 time period.

```

# Ongava historical PPT data

Import Ongava data from Dropbox\
Note that these data are from Stuart Crawford via Rolf Becker. Stuart's email suggests that more data may be available from the Davis Weatherlink app. I didn't find any publicly available data from the app, but can look into this.

```{r}
# Import Ongave rainguage data using the corrected Dropbox URL for direct download
url <- "https://www.dropbox.com/scl/fi/8r6u8lwu2llrow02wmtxq/OngavaRainData.csv?rlkey=0jycso3d15qudsoazzogwmcsg&dl=1"

# Read the CSV file from the corrected URL
Ongava_rainfall <- read.csv(url)

# Join the annual totals back to the original data
precip_data_2010.2019 <- precip_data_2010.2019 |>
    left_join(Ongava_rainfall, by = "rain_year")
```

### Ongava vs WorldClim data

```{r}
# Plot Ongava measured versus WorldClim rain_year data
precip_data_2010.2019$rain_year <- 
  factor(precip_data_2010.2019$rain_year, levels = unique(precip_data_2010.2019$rain_year))

# Remove monthly data
annual_rainfall_2010.2019 <- precip_data_2010.2019 |>
    filter(Month == 1) |>
    filter(rain_year != "2009-2010") |> # remove 2009-2010 since no 2009 data
    filter(rain_year != "2019-2020") |> # remove 2019-2020 since no 2020 data
    select(rain_year, Annual_PPT, OngavaMeasuredPPT)

# calculate mean rain_year PPT for both data sources
mean_WC_rainyr_ppt2010.2019 <- mean(
  annual_rainfall_2010.2019$Annual_PPT, na.rm = TRUE)
mean_Ong_rainyr_ppt2010.2019 <- mean(
  annual_rainfall_2010.2019$OngavaMeasuredPPT, na.rm = TRUE)

# plot the rainfall over time
rainfall_comparison <- ggplot(annual_rainfall_2010.2019, aes(x = rain_year)) +
  geom_line(aes(y = Annual_PPT, group = 1, 
                color = "WorldClim2.1", linetype = "WorldClim2.1"), 
            linewidth = 1) +
  geom_line(aes(y = OngavaMeasuredPPT, group = 1, 
                color = "Ongava rain gauge", linetype = "Ongava rain gauge"),
            linewidth = 1) +
  scale_color_manual(name = "Legend", values = c("WorldClim2.1" = "black", 
            "Ongava rain gauge" = "red")) +
  scale_linetype_manual(name = "Legend", values = c("WorldClim2.1" = "solid",              "Ongava rain gauge" = "dashed")) +
  geom_hline(yintercept = mean_Ong_rainyr_ppt2010.2019, color = "red", 
             linetype = "dotted", size = 1) + # line for mean rainfall PPT 
  geom_hline(yintercept = mean_WC_rainyr_ppt2010.2019, color = "black", 
             linetype = "dotted", size = 1) + # line for mean rainfall PPT 
  labs(x = "Rain Year", y = "Precipitation (mm)", 
       title = "") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "top")
rainfall_comparison

ggsave(filename = here("figs", "Ongava_vs_WorldCLim_annualPPT.jpg"),
       plot = rainfall_comparison,
       device = "jpeg",
       width = 8, height = 6, dpi = 300)
```

# PPT Addition Calculations

```{r}
# calculate monthly PPT addition (mm) if we add the mean rainfall on top of ambient rainfall

# calculate area of the plot and save as a variable
plotarea <- 3.14159*0.5^2
# a 1 mm PPT event is equivalent to application of 0.785 L added to each plot

# calculate monthly PPT addition (L) and their distribution for each plot
# These calculations are based on mean monthly PPT (assuming adding the eqivalent of monthly rainfall).
# PPT_events_per_month is the number of times that rainfall is added, which is 4 by default (for weekly rainfall) but 1 if less than 10 L are added per month. L_per_event is the amount that should be added to each plot for each rainfall event.
monthly_PPT_additions <- monthly_precip_1960.2021 |>
    mutate(monthlyPPTadd_L_per_plot = mean_monthly_PPT * plotarea,
          PPT_events_per_month = ifelse(near(monthlyPPTadd_L_per_plot, 0), 0,
                                ifelse(monthlyPPTadd_L_per_plot < 10, 1, 4)),
          L_per_event = ifelse(near(monthlyPPTadd_L_per_plot, 0), 0, 
                        monthlyPPTadd_L_per_plot / PPT_events_per_month)
          ) |>
    select(Month,mean_monthly_PPT,monthlyPPTadd_L_per_plot,
           PPT_events_per_month,L_per_event)

write.csv(monthly_PPT_additions, file = here(
  "output", "monthly_PPT_additions.csv"), row.names = FALSE)

```

# Ongava Contemporary Data

### 2024-5 Monthly means vs historical WorldClim

```{r}
# compare monthly PPT from WorldClim2.1 data with rainguage data collected at Ongava during the dung decay experiment

# Load the Ongava rainfall dataset
ongava_ppt_2024_2025 <- read_csv("https://www.dropbox.com/scl/fi/cb3g6no7ddaohm46lxcei/OngavaRain2024_2025.csv?rlkey=ulh5lpyazi0hxdhdv6szdh5nt&dl=1")

# Convert date column to Date format
ongava_ppt_2024_2025 <- ongava_ppt_2024_2025 |>
  mutate(Date = mdy(Date),  # Use mdy() for "month/day/year" format
         Year = year(Date),
         Month = month(Date))

# Aggregate daily rainfall to monthly totals
monthly_rainfall_2024_2025 <- ongava_ppt_2024_2025 |>
  group_by(Year, Month) |>
  summarize(total_PPT = sum(CampusPPT, na.rm = TRUE), .groups = "drop")

# Create the original plot
monthly_rainfall_plot <- ggplot(monthly_precip_2010.2019, aes(x = Month)) +
  geom_line(aes(y = mean_monthly_PPT), linewidth = 1) +
  geom_errorbar(aes(ymin = mean_monthly_PPT - se_monthly_PPT, 
                    ymax = mean_monthly_PPT + se_monthly_PPT), width = 0.2) +
  scale_x_continuous(breaks = 1:12) +
  scale_y_continuous(expand = c(0, 0)) +
  labs(x = "Month", y = "Precipitation (mm)", 
       title = "Ongava WorldClim2.1 2009-2019 vs. Campus Rainfall") +
  theme_classic()

# Add observed rainfall as individual points with different symbols for each year
monthly_rainfall_plot <- monthly_rainfall_plot +
  geom_point(data = monthly_rainfall_2024_2025, 
             aes(x = Month, y = total_PPT, shape = factor(Year)), size = 3) +
  scale_shape_manual(name = "Year", values = seq(0, 9))  # Unique symbols for years

# Add red text annotation in the middle of the plot
monthly_rainfall_plot <- monthly_rainfall_plot +
  annotate("text", x = 6.5, y = max(monthly_precip_2010.2019$mean_monthly_PPT, na.rm = TRUE) * 0.5, 
           label = "data through 2025-04-03", color = "red", size = 4)

# Display the plot
print(monthly_rainfall_plot)

# Save the updated plot
ggsave(filename = here("figs", "WorldClim_vs_actual_monthly_rainfall.jpg"),
       plot = monthly_rainfall_plot,
       device = "jpeg",
       width = 8, height = 6, dpi = 300)
```
