---
title: "OngavaDecomp Kestrel Data"
author: "Heather Throop"
format: html
editor: visual
editor_options: 
  chunk_output_type: console
---

Data compilation of Kestrel data from Ongava for the decomposition experiment (2024-2026)

# Overview

This script reads in csv files from individual Kestrel loggers, merges the files, and synthesizes the data.

Created January 24, 2025 Heather Throop\
Updated with new data 2025-03-19, 2025-04-09, 2025-12-06

Raw Kestrel files are stored in the "L0" folder for this project on Dropbox. This code pulls data from the Dropbox folder and writes figures and csv files to Dropbox as mapped with a config folder.

```{r}
#| label: load-packages
library(here) # v. 0.1
library(stringr) # v. 1.2.0
library(purrr) # v. 0.2.3
library(tidyverse)
library(scales)
library(lubridate)
```

Display information about package versions in use when this code was run.

```{r}
#| label: session-info
sessionInfo()
```

```{r}
#| label: config-setup

# This script uses a config.yml file to define the path for the DERT Data & Metadata folder (in Dropbox) where Heather has the data stored. This can be changed to your own path as needed or you can read in data from elsewhere, but you will need to update the read-in-files steps. 
DERT_Archives <- config::get()
```

# L0: Read, Clean, Combine

### List Kestrel Files

Note that the "more columns than column names" error is likely to happen. Find the offending file (use the last line of code in this chunk to do so) and delete extra columns to the right of the data.\
\
Much of this code is based on the commentary from: https://aosmith.rbind.io/2017/12/31/many-datasets/

```{r}
# list all files to read in
### Note that this reference from Heather's laptop and may need to be changed for a different machine
allfiles <- list.files(
  path = file.path(DERT_Archives, "OngavaDungDecay", "MicroclimateData", "L0"),  
  pattern = "\\.csv$",  # Ensure proper regex for .csv files  
  full.names = TRUE  
)

# Create a function to read and extract loggerID
read_fun = function(path) {
  # Read in the file without assigning column names yet
  test = read.csv(path, skip = 5, header = FALSE)
  
  # Keep only the first 6 columns before assigning names
  test = test[, 1:6]
  
  # Assign column names
  colnames(test) = c("datetime", "temperature", "RH", "heatindex", "dewpoint", "datatype")
  
  # Extract loggerID from the file name
  allnames = str_split(path, pattern = "/", simplify = TRUE)
  test$loggerID = str_extract(allnames[, ncol(allnames)], pattern = "2[0-9]{6}") # 7-digit logger ID starting with 2 and then 6 digits that are between 0 and 9
  
  # Coerce columns to numeric
  test$RH <- as.numeric(as.character(test$RH))
  test$heatindex <- as.numeric(as.character(test$heatindex))
  test$dewpoint <- as.numeric(as.character(test$dewpoint))
  test$temperature <- as.numeric(as.character(test$temperature))
  test$loggerID <- as.numeric(as.character(test$loggerID))
  
  test
}

# Line below is useful for testing if logger name is being assigned
# change the number in brackets to check show the top section of each file
read_fun(allfiles[35]) # use to test the function on an individual file
```

### Read in and Combine Files

```{r}
#| label: combine-kestrel-files
combined_kestrel_notcleaned = map_dfr(allfiles, read_fun) # combine all the files

# check that we have the right number of loggers in the file
combined_kestrel_notcleaned |>
  group_by(loggerID) |> 
  summarize(n = n())

# Add in microsite and block information
## pulls in file with loggerID information ("KestrelLoggerAssignments")
df_loggers <- read.csv(file.path(DERT_Archives, "OngavaDungDecay",
            "MicroclimateData", "KestrelLoggerAssignments.csv"))

## joins the dataframes
combined_kestrel_notcleaned <- combined_kestrel_notcleaned |>
  left_join(
    df_loggers |> 
      select(loggerID, Block, Microsite, WaterTrt, startdate),
    by = "loggerID"
  )

#save the output as a .csv file - this is the combined L0 data that has not yet been cleaned
write_csv(combined_kestrel_notcleaned,
  file.path(DERT_Archives, "OngavaDungDecay","MicroclimateData",
            "L1", "Ongava_kestrel_L0_combined.csv"))
```

## L1

### Time formatting

Check that the time column is populated *for all loggers*. If not, the format might be incorrect - sometimes Kestels are set to send the datetime output in different formats.

```{r}
#| label: Fix-datetime-formatting 

# make a new file for messing around with date-time formats
combined_kestrel_cleaning <- combined_kestrel_notcleaned 

# Initialize the 'time' column
combined_kestrel_cleaning$time <- NA

# Try parsing with flexible parse_date_time function
combined_kestrel_cleaning$time <- parse_date_time(
  combined_kestrel_cleaning$datetime, 
  orders = c("mdy HM", "ymd HMS p"),  # List of possible formats
  quiet = FALSE  # This will show warnings and errors if parsing fails
)

# Check if there are still NA values after parsing
if (any(is.na(combined_kestrel_cleaning$time))) {
  cat("Some entries could not be parsed!\n")
  # Print rows where the time column is still NA
  print(combined_kestrel_cleaning[is.na(combined_kestrel_cleaning$time),])
}

# Remove rows with NA time values (if any remain after conversion)
combined_kestrel_cleaning <- combined_kestrel_cleaning[!is.na(combined_kestrel_cleaning$time), ]

# Ensure the time is in 24-hour format (POSIXct will handle it correctly)
combined_kestrel_cleaning$time <- 
  as.POSIXct(combined_kestrel_cleaning$time, format = "%Y-%m-%d %H:%M:%S")

# Round down to the nearest minute (sets seconds to :00)
# Accounts for loggers that oddly start logging at :30
combined_kestrel_cleaning$time <- 
  floor_date(combined_kestrel_cleaning$time, unit = "minute")

# remove any data before the deployment date and after the collection date
# some loggers were added part way through the experiment so startdate varies among loggers
combined_kestrel_cleaning$startdate <- as.POSIXct(mdy(combined_kestrel_cleaning$startdate))

L1_kestrel_data <- combined_kestrel_cleaning |>
  filter(
    !is.na(startdate),                       # remove rows with missing startdate
    as.Date(time) >= as.Date(startdate),     # compare dates, ignore time-of-day
    time <= as.POSIXct("2025-10-08 23:59:59") # fixed end datetime
  )

# At least one logger was logging every 10 minutes. Keep only logs on the hour
L1_kestrel_data <- L1_kestrel_data |> 
  filter(minute(time) == 0,
         second(time) == 0)

# remove any repetitive rows where there are multiple entries for the same time, date, and loggerID
L1_kestrel_data <- L1_kestrel_data |>
  distinct(time, loggerID, .keep_all = TRUE)

# Check that all loggers are still present
L1_kestrel_data |>
  group_by(loggerID) |> 
  summarize(n = n())

write_csv(L1_kestrel_data,
  file.path(DERT_Archives, "OngavaDungDecay","MicroclimateData",
            "L1", "Ongava_kestrel_L1.csv"))

# check the data carefully to see if any additional cleaning is needed!
```

### Individual Logger Plots

Below are separate plots for each logger (grouped by treatment). These are intended to check the data (e.g., gaps in logs for different loggers and correct number of loggers per site).

#### RH

```{r}
#| label: RH-plot

p1 <- L1_kestrel_data |>
  ggplot(aes(x=time, y=RH, color = WaterTrt)) + 
    labs(title = " ") +
    geom_line() +
    xlab(" ") +
    ylab("RH (%)") +
    scale_x_datetime(date_breaks = "1 week", labels = date_format("%m-%d-%y")) +
    facet_grid(loggerID ~ .)  +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
p1

# Save the plot as a JPG in the "figs" folder
ggsave(
  filename = file.path(DERT_Archives, "OngavaDungDecay", "MicroclimateData",
                       "figs", "RH_all_loggers.jpg"),
  plot = p1,
  width = 8,
  height = 6,
  dpi = 300
)
```

#### Temperature

```{r}

p2 <- L1_kestrel_data |>
  ggplot(aes(x=time, y=temperature, color = WaterTrt, group = 1)) + 
    labs(title = " ") +
    geom_line() +
    xlab(" ") +
    ylab("Temperature (C)") +
    scale_x_datetime(date_breaks = "1 week", labels = date_format("%m-%d-%y")) +
    facet_grid(loggerID ~ .) +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
p2

# Save the plot as a JPG in the "figs" folder
ggsave(
  filename = file.path(DERT_Archives, "OngavaDungDecay", "MicroclimateData",
                       "figs", "Temp_all_loggers.jpg"),
  plot = p2,
  width = 8,
  height = 6,
  dpi = 300
)
```

# L2: Treatment Means

*Check that the tibble output after running this chunk looks reasonable. N should equal the expected number of loggers. There should be values for SD and SE, assuming that there is more than one logger for each time point.*

```{r}
#| label: hourly-treatment-means
# create df that condenses the multiple loggers per microsite and water treatment down to one mean value per hour

L2_hourlymeans_df <- L1_kestrel_data |>
  group_by(Microsite, WaterTrt, time) |>
  summarize(
               N    = sum(!is.na(RH)),
               mean_T = mean(temperature, na.rm=TRUE),
               sd_T   = sd(temperature, na.rm=TRUE),
               se_T   = sd_T / sqrt(N),
               mean_RH = mean(RH, na.rm=TRUE),
               sd_RH   = sd(RH, na.rm=TRUE),
               se_RH   = sd_RH / sqrt(N)
               )

```

## Plots of hourly means

### RH

```{r}
#| label: RH-by-treatment

# plots of hourly means by treatment - RH
p3 <- L2_hourlymeans_df |>
  ggplot(aes(x=time, y=mean_RH, color = WaterTrt, linetype = WaterTrt)) + 
  geom_line() +
  theme_bw() +
  ylab("Relative Humidity (%)") +
  xlab("") +
  facet_grid(Microsite ~ WaterTrt) +
  scale_color_manual(values = c("Addition" = "blue", "Ctrl" = "red")) +  
  scale_linetype_manual(values = c("Addition" = "solid", "Ctrl" = "solid"))  
p3

# Save the plot as a JPG in the "figs" folder
ggsave(
  filename = file.path(DERT_Archives, "OngavaDungDecay", "MicroclimateData",
                       "figs", "RH_by_trt.jpg"),
  plot = p3,
  width = 8,
  height = 6,
  dpi = 300
)

```

### Temperature

```{r}
#| label: temperature-by-treatment

# plots of hourly means by treatment - temperature
p4 <- L2_hourlymeans_df |>
ggplot(aes(x=time, y=mean_T, color = WaterTrt, linetype = WaterTrt)) + 
  geom_line() +
  theme_bw() +
  ylab("Temperature (C)") +
  xlab("") +
  facet_grid(Microsite ~ WaterTrt) +
  scale_color_manual(values = c("Addition" = "blue", "Ctrl" = "red")) +  
  scale_linetype_manual(values = c("Addition" = "solid", "Ctrl" = "solid"))
p4

# Save the plot as a JPG in the "figs" folder
ggsave(
  filename = file.path(DERT_Archives, "OngavaDungDecay", "MicroclimateData",
                       "figs", "Temp_by_trt.jpg"),
  plot = p4,
  width = 8,
  height = 6,
  dpi = 300
)

```
